Hydrous melting and partitioning in and above the mantle transition zone: insights from water-rich MgO-SiO$_2$-H$_2$O experiments

### 2015-07-10 MSH.tex
Checklist
Update

\begin{figure}[h!]
  \centering
      %\includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
  \caption{The 18/11 octahedral assembly design used in this study.}
  \label{fig:assembly}
\end{figure}

@@

Partitioning of hydrogen between mineral and melt phases can be described using a partition coefficient D:
\begin{equation}
  D^{a/b}_{H_2O} = c^a_{H_2O} / c^b_{H_2O}
\end{equation}
where $c^i_{H_2O}$ is the concentration of H$_2$O (in wt\%) in phase $i$ \citep{KB2006}. In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA. In the case of melts, which are unquenchable, $c^{melt}_{H_2O}$ in partitioning studies has most commonly been estimated by mass balance or by assuming that the deficit in microprobe totals is due to hydrogen. In two studies on Fe-free wadsleyite, \cite{DDFK2005} and \cite{LSOK2011} present estimates of water content in melts calculated from microprobe deficits. A similar study on ringwoodite was undertaken by \cite{OMY2000}. The T-X$_{H_2O}$ dependencies of the melts in equilibrium with wadsleyite or ringwoodite are strikingly different (Figure \ref{fig:partitioning}). For example, melt in equilibrium with wadsleyite at 15 GPa and 1400 $^{\circ}$C was estimated to contain 10.6-13.3 wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite at 20 GPa and 1400 $^{\circ}$C suggest water contents of 57-66 wt\% H$_2$O. Modelled Mg$_2$SiO$_4$ melting temperatures increase by only 40 K between 15 and 20 GPa, and entropies of melting at these two pressures are not very different (92.4 vs. 97 J/K/mol), so unless the thermodynamics of mixing in the liquid change markedly between these pressures, estimates of $c^{melt}_{H_2O}$ for the ringwoodite or wadsleyite studies (or both) must be somewhat inaccurate. We note that two of the three values provided by \cite{OMY2000} fall above the temperatures indicating that H$_2$O is the only H-bearing species in the melt \citep{SS1985}, and that the melting temperatures of wadsleyite + 20.4 wt\% H$_2$O (X$_{H_2O}$=0.4 in Figure \ref{fig:partitioning}) in the wadsleyite studies fall below that of the same composition at 5.5 GPa \citep{Inoue1994}. Both of these interpretations appear unlikely.

@@

It is also possible to create a thermodynamic model for incorporation of H$_2$O into wadsleyite and ringwoodite, given some knowledge of the substitution mechanism. Hydrogen in wadsleyite and ringwoodite is incorporated as proton pairs substituting for Mg (Mg$^{2+}$ $\leftarrow$ 2H$^{+}$) \citep{Smyth1987, SHFJLM2003}. Ringwoodite potentially also exhibits a substitution involving Si (Si$^{4+}$ $\leftarrow$ Mg$^{2+}$ + 2H$^{+}$) \citep{KKMO2000}. Both of these reactions may be written in the form 
\begin{equation}
H_2O + O = (OH)_2
\end{equation}
which leads to the equilibrium relation 
\begin{eqnarray}
\Delta G = -R T \ln \left( \frac{a_{(OH)_2}}{a_{H_2O}a_{O}} \right), \\
= \Delta H - (T-T_0) \Delta S + (P-P_0) \Delta V
\end{eqnarray}
where $\Delta G$ is the Gibbs energy associated with the reaction. The second equation is a linearised form of the expression describing the change in free energy with pressure and temperature. The first equation can be simplified by making the approximation that $a_O$ is constant and that the weight percentage of H$_2$O in the solid $c_{H_2O}$ is proportional to the activity of (OH)$_2$. Both of these assumptions are probably reasonable at low values of $c_{H_2O}$. Rearranging now produces the expression:
\begin{equation}
c_{H_2O} \propto a_{H_2O} \exp{ \left( -\frac{\Delta H - (T-T_0) \Delta S + (P-P_0) \Delta V}{R T} \right) }
\end{equation}
Further simplifications can be made by taking $T_0 = 0$ K, by realising that $P_0$ can be neglected at high pressures, and by incorporating the factor $\Delta S/R$ into the exponential prefactor. Finally, it is recognised that the pressure and temperature dependence of the Gibbs free energy of water $G_{H_2O}$ is highly nonlinear. For this reason, we split the Gibbs free energy change of the anhydrous and hydrous solid from that of the liquid, yielding the equation \citep{KB2006}:
\begin{eqnarray}
c_{H_2O} = A a_{H_2O} f_{H_2O} \exp{ \left( -\frac{\Delta H + P \Delta V^{solid}}{R T} \right) }, \\
f_{H_2O} = \exp \left(\frac{\Delta G_{H_2O}}{RT} \right)
\end{eqnarray}

We calculate the fugacity of pure water $f_{H_2O}$ from the equation of state of \citep{PS1995}. We can now fit the values of A, $\Delta H$ and $\Delta V$ to the experimental data on wadsleyite and ringwoodite using our melt model. The pressure data of \cite{DDFK2005} at 1200 $^{\circ}$C and 14--17 GPa are consistent with a negligable pressure dependence on water content of wadsleyite, implying that $\Delta V \sim 10^{-5}$ m$^3$/mol, similar or slightly lower than estimates for olivine \citep[1.00 -- 1.06 $\cdot 10^{-5}$ m$^3$/mol][]{KKR1996, ZGK2004, MDAR2006}. A single data point at 18 GPa has much lower water contents, which would require $\Delta V \sim 1.25 \cdot 10^{-5}$ m$^3$/mol and would produce an almost linear curve which produces large misfits at 16 and 17 GPa. We therefore favour the lower estimate of $\Delta V$, and assume the same value for ringwoodite. Other estimated values are shown in Table \ref{table:partitioning}.

@@

These models allow us to predict the partition coefficient of water between wadsleyite and ringwoodite at the 520 km discontinuity ($\sim$18 GPa). The coefficent increases gradually with temperature between 0.6 at 1000 $^{\circ}$C to almost 1.0 at 2000 $^{\circ}$C. One implication is that repartitioning of water during secular cooling of the Earth is likely to be much more minor (and in the opposite direction) to that proposed by \cite{DDFK2005}. The second implication is that water saturated peridotite in the lower mantle will partially melt during upwelling, which could explain observations of low velocity layers at ca. 520 km depth \citep[e.g.][]{JDH2010}.

>>>>

\begin{figure}[h!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/assembly}
  \caption{The 18/11 octahedral assembly design used in this study.}
  \label{fig:assembly}
\end{figure}

@@

Partitioning of hydrogen between mineral and melt phases can be described using a partition coefficient D:
\begin{equation}
  D^{a/b}_{H_2O} = c^a_{H_2O} / c^b_{H_2O}
\end{equation}
where $c^i_{H_2O}$ is the concentration of H$_2$O (in wt\%) in phase $i$ \citep{KB2006}. In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA. In the case of melts, which are unquenchable, $c^{melt}_{H_2O}$ in partitioning studies has most commonly been estimated by mass balance or by assuming that the deficit in microprobe totals is due to hydrogen. In two studies on Fe-free wadsleyite, \cite{DDFK2005} and \cite{LSOK2011} present estimates of water content in melts calculated from microprobe deficits. A similar study on ringwoodite was undertaken by \cite{OMY2000}. The T-X$_{H_2O}$ dependencies of the melts in equilibrium with wadsleyite or ringwoodite are strikingly different (Figure \ref{fig:partitioning}). For example, melt in equilibrium with wadsleyite at 15 GPa and 1400 $^{\circ}$C was estimated to contain 10.6-13.3 wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite at 20 GPa and 1400 $^{\circ}$C suggest water contents of 57-66 wt\% H$_2$O. Modelled Mg$_2$SiO$_4$ melting temperatures increase by only 40 K between 15 and 20 GPa, and entropies of melting at these two pressures are not very different (92.4 vs. 97 J/K/mol), so unless the thermodynamics of mixing in the liquid change markedly between these pressures, estimates of $c^{melt}_{H_2O}$ for the ringwoodite or wadsleyite studies (or both) must be somewhat inaccurate. We note that two of the three melt compositions in equilibrium with ringwoodite fall above the temperatures indicating that hydrogen exists only as molecular H$_2$O \citep{SS1985}. In addition, the calculated liquidus temperatures in the wadsleyite studies at 15 GPa are lower than those reported at 5.5 GPa \citep{Inoue1994}. Both of these observations appear highly unlikely.

@@

\begin{figure}[ht!]
  \centering
      %\includegraphics[width=0.8\textwidth]{figures/partitioning}
  \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}. Inset: Modelled partition coefficients between wadsleyite, ringwoodite and melt at 18 GPa.}
  \label{fig:partitioning}
\end{figure}

We now create a thermodynamic model for incorporation of H$_2$O into wadsleyite and ringwoodite. Hydrogen in wadsleyite and ringwoodite is incorporated as proton pairs substituting for Mg (Mg$^{2+}$ $\leftarrow$ 2H$^{+}$) \citep{Smyth1987, SHFJLM2003}. Ringwoodite potentially also exhibits a substitution involving Si (Si$^{4+}$ $\leftarrow$ Mg$^{2+}$ + 2H$^{+}$) \citep{KKMO2000}. Both of these reactions may be written in the form 
\begin{equation}
H_2O + O = (OH)_2
\end{equation}
which leads to the equilibrium relation 
\begin{eqnarray}
\Delta G = -R T \ln \left( \frac{a_{(OH)_2}}{a_{H_2O}a_{O}} \right), \\
= \Delta H - (T-T_0) \Delta S + (P-P_0) \Delta V
\end{eqnarray}
where $\Delta G$ is the Gibbs energy associated with the reaction. The second equation is a linearised form of the expression describing the change in free energy with pressure and temperature. The first equation can be simplified by making the approximation that $a_O$ is constant and that the weight percentage of H$_2$O in the solid $c_{H_2O}$ is proportional to the activity of (OH)$_2$. Both of these assumptions are probably reasonable at low values of $c_{H_2O}$. Rearranging now produces the expression:
\begin{equation}
c_{H_2O} \propto a_{H_2O} \exp{ \left( -\frac{\Delta H - (T-T_0) \Delta S + (P-P_0) \Delta V}{R T} \right) }
\end{equation}
Further simplifications can be made by taking $T_0 = 0$ K, by realising that $P_0$ can be neglected at high pressures, and by incorporating the factor $\Delta S/R$ into the exponential prefactor. Finally, it is recognised that the pressure and temperature dependence of the Gibbs free energy of water $G_{H_2O}$ is highly nonlinear. For this reason, we split the Gibbs free energy change of the anhydrous and hydrous solid from that of the liquid, yielding the equation \citep{KB2006}:
\begin{eqnarray}
c_{H_2O} = A a_{H_2O} f_{H_2O} \exp{ \left( -\frac{\Delta H + P \Delta V^{solid}}{R T} \right) }, \\
f_{H_2O} = \exp \left(\frac{\Delta G_{H_2O}}{RT} \right)
\end{eqnarray}

We calculate the fugacity of pure water $f_{H_2O}$ from the equation of state of \citep{PS1995}. We can now fit the values of A, $\Delta H$ and $\Delta V$ to the experimental data on wadsleyite and ringwoodite using our melt model. The pressure data of \cite{DDFK2005} at 1200 $^{\circ}$C and 14--17 GPa are consistent with a negligable pressure dependence on water content of wadsleyite, implying that $\Delta V \sim 10^{-5}$ m$^3$/mol, similar or slightly lower than estimates for olivine \citep[1.00 -- 1.06 $\cdot 10^{-5}$ m$^3$/mol][]{KKR1996, ZGK2004, MDAR2006}. A single data point at 18 GPa has much lower water contents, which would require $\Delta V \sim 1.25 \cdot 10^{-5}$ m$^3$/mol. This value is inconsistent with the large water contents at 16 and 17 GPa, and is probably unreasonable given the reported values for olivine. We therefore favour the lower estimate of $\Delta V$, and assume the same value for ringwoodite. The model values are shown in Table \ref{table:partitioning}.

@@

These models allow us to predict the partition coefficients of water between wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa). $DD^{wad/ring}_{H_2O}$ increases from 0.6 at 1000 $^{\circ}$C to almost 1.0 at 2000 $^{\circ}$C. That these values are <1 indicate that water saturated peridotite in the lower mantle will partially melt during upwelling, which could explain observations of low velocity layers at ca. 520 km depth \citep[e.g.][]{JDH2010}. The small temperature dependence implies that repartitioning of water during secular cooling of the Earth is likely to be more minor (and in the opposite direction) to that proposed by \cite{DDFK2005}.
